General definitions

### 2022-07-10 2021/algebra.tex
putting old stuff into its own folder

@@

We now turn to quantization, and our phase space will be, for simplicity, the algebraic loop space $V[z,z^{-1}]$ of a symplectic complex vector space $V$.
Following the case of mechanics, we might ask for a deformation of $\cO(V[z,z^{-1}])$ into an associative algebra.
A subtlety here is that $\cO(V[z,z^{-1}])$ does not possess a Poisson bracket, 
due to the  infinite-dimensionality of $V[z,z^{-1}]$.
The naive formulas from functions on a finite-dimensional symplectic vector space only make sense on a subset of observables here.
One workaround is to replace $V[z,z^{-1}]$ with $V((z))$, the $V$-valued Laurent series,
which we view as ``formal loops in $V$.''
There is manifestly an inclusion $V[z,z^{-1}] \hookrightarrow V((z))$ and hence an algebra map $\cO(V((z))) \to \cO(V[z,z^{-1}])$
where 
\[
\cO(V((z))) = \Sym(V^* [z,z^{-1}]).
\]
On this subalgebra, the naive Poisson bracket makes sense: it is given by extending the nondegenerate skew-symmetric pairing $\omega^* \otimes \Res$ on $V^* [z,z^{-1}]$.
Hence for $\cO(V((z)))$, we can ask for a deformation quantization,
and there is a nice answer: 
the Heisenberg commutation relations determine a Weyl algebra $W_{V}$ for this formal loop space.
As a vector space $W_V$ is isomorphic to $\cO(V((z)))$, 
just as the simplest Weyl algebra is isomorphic to $\CC[p,q]$ as a vector space,
but the product structure is modified by quantization.
We expect, by analogy with the simplest case, that any quantization of a bigger algebra (e.g., for observables of the holomorphic loop space) contains this quantization as its algebraic skeleton.

The subspace of ``contractible formal loops" $V[[z]]$ inside $V((z))$ produces a module for $\cO(V((z)))$ given by $\cO(V[[z]]$. 
(This module is in fact a quotient algebra.)
One can quantize this module to a module ${\rm Vac}_V$ for $W_V$,
where as a vector space, it is still $\cO(V[[z]])$.
This module structure means there is a map of algebras $W_V \to \End({\rm Vac}_V)$,
so we have a natural ``Hilbert space'' or Fock module for~$W_V$.

It is a remarkable fact, suggested by physicists, that there is also a map
\[
Y: {\rm Vac}_V \to \End({\rm Vac}_V)[[z, z^{-1}]],
\]
known as the {\em vertex operator} or {\em state-field correspondence},
which is closely related to the $W_V$-action.
The intuition behind $Y$ is that for any map $\phi: \CC^\times \to V$ and for any point $w \in \CC^\times$, 
there is a restriction of $\phi$ to a little disk around $w$.
That is, there is a restriction map $r_w: V[z,z^{-1}] \to V[[t]]$, where $t$ is the local coordinate $t = z-w$.
Hence there is a $w$-dependent map $\cO(V[[t]]) \to \cO(V[z,z^{-1}])$ by pulling back a function along $r_w$;
this map determines a $w$-dependent action of $\cO(V[[t]])$ on $\cO(V[[z]])$.
One can, in essence, quantize this map, by trying to extend the formulas of the $W_V$-action,
and this leads to the map~$Y$.
If one axiomatizes the behavior of $Y$, one is led to the notion of a vertex algebra.

@@

When we try to quantize, we run into an interesting phenomenon:
the Hamiltonian action $H_\rho^{\rm loc}$ does not extend as a Lie algebra map into the Weyl algebra $W_V$,
although it makes sense as a linear map.
The failure to respect the brackets determines, however, a cocycle on the loop algebra
by
\[
\alpha_V( x \otimes f(z) , y \otimes g(z)) = \Tr_V(xy) \Res(f \, \partial_z g).
\]
\owen{Check conventions.}
There is thus a central extension of the loop algebra 
\[
\CC c \to \widehat{\fg} \to \fg[z,z^{-1}]
\]
so that 
\[
[x \otimes f, y \otimes g] = [x,y] \otimes fg + \alpha_V( x \otimes f , y \otimes g) c,
\]
and there is a Lie algebra map
\[
\widehat{H}^{\rm loc}_\rho: \widehat{\fg} \to W_V,
\]
sometimes called {\em free field realization} of the affine Lie algebra~$\widehat{\fg}$.
This relationship --- and its extension to a map between the vertex algebras associated to $\widehat{\fg}$ and $V$ --- is a modest \owen{better adjective?} extract of the rich dialogue that has developed between complex analysis, representation theory, and the physics of chiral conformal field theory.
(To get Riemann surfaces into the game, we will need a richer framework than ordinary algebra.)

@@

There are natural central extensions of this sphere algebra as {\em $L_\infty$ algebras},
parallel to the affine Lie algebras appearing as central extensions of the loop algebra.
An $L_\infty$ algebra is a generalization of a Lie algebra;
it is graded vector space $L$ equipped with a countable collection of ``higher brackets'' $[-]_k: L^{\otimes k} \to L$ that satisfy a system of equations generalizing the Jacobi condition for a standard Lie algebra.
For an overview (and actual definition!) we recommend \owen{add citations}.

In our setting of $\fg^\bullet_d$, Faonte-Hennion-Kapranov introduced central extensions that only affect the $d$-bracket.
For any $\fg$-invariant degree $d+1$ polynomial  $\theta \in \Sym^{d+1}(\fg^*)^\fg$,  the FHK  cocycle is
\[
\label{fhk cocycle}
\begin{array}{cccc}
\theta_{\rm FHK} : & (A_d \tensor \fg)^{\tensor (d+1)} & \to & \CC\\ 
& a_0 \otimes \cdots \otimes a_d & \mapsto & \Res_{z=0} \theta(a_0,\partial a_1,\ldots,\partial a_d)
\end{array}.
\]
This cocycle has cohomological degree $2$ and so determines an unshifted central extension as $L_\infty$ algebras
\[
\CC c \to \widetilde{\fg}^\bullet_{d, \theta} \to \fg^\bullet_d ,
\]
so that the usual bracket (the 2-bracket) of $\fg^\bullet_d$ is unchanged but there is now a nontrivial $d$-bracket defined by~$\theta_{\rm FHK}$.

>>>>

\begin{rmk}
We have only spoken of symplectic manifolds, and emphasized vector spaces,
but fermions play a crucial role in quantum mechanics too.
Here the symmetric algebra on a symplectic vector space (as classical observables) 
is replaced by an exterior algebra on an inner product space,
which can be seen as a symmetric algebra on an {\em odd} vector space
(i.e., a super vector space with purely odd component).
The deformation quantization of the exterior algebra is then a Clifford algebra.
We will set fermions aside for the remainder of this section,
but the discussion below extends easily to them and fermions appear naturally in the study of holomorphic field theories.
\end{rmk}

@@

We now turn to quantization, and our phase space will be, for simplicity, the algebraic loop space $V[z,z^{-1}]$ of a symplectic complex vector space $V$.
Following the case of mechanics, we might ask for a deformation of $\cO(V[z,z^{-1}])$ into an associative algebra.
A subtlety here is that $\cO(V[z,z^{-1}])$ does not possess a Poisson bracket, 
due to the  infinite-dimensionality of $V[z,z^{-1}]$.
The naive formulas from functions on a finite-dimensional symplectic vector space only make sense on a subset of observables here.
One workaround is to replace $V[z,z^{-1}]$ with $V((z))$, the $V$-valued Laurent series,
which we view as ``formal loops in $V$.''
There is manifestly an inclusion $V[z,z^{-1}] \hookrightarrow V((z))$ and hence an algebra map $\cO(V((z))) \to \cO(V[z,z^{-1}])$
where 
\[
\cO(V((z))) = \Sym(V^* [z,z^{-1}]).
\]
On this subalgebra, the naive Poisson bracket makes sense: it is given by extending the nondegenerate skew-symmetric pairing $\omega^* \otimes \Res$ on $V^* [z,z^{-1}]$.
Hence for $\cO(V((z)))$, we can ask for a deformation quantization,
and there is a nice answer: 
the Heisenberg commutation relations determine a Weyl algebra $W[V]$ for this formal loop space.
As a vector space $W[V]$ is isomorphic to $\cO(V((z)))$, 
just as the simplest Weyl algebra is isomorphic to $\CC[p,q]$ as a vector space,
but the product structure is modified by quantization.
We expect, by analogy with the simplest case, that any quantization of a bigger algebra (e.g., for observables of the holomorphic loop space) contains this quantization as its algebraic skeleton.

The subspace of ``contractible formal loops" $V[[z]]$ inside $V((z))$ produces a module for $\cO(V((z)))$ given by $\cO(V[[z]]$. 
(This module is in fact a quotient algebra.)
One can quantize this module to a module ${\rm Vac}[V]$ for $W[V]$,
where as a vector space, it is still $\cO(V[[z]])$.
This module structure means there is a map of algebras $W[V] \to \End({\rm Vac}[V])$,
so we have a natural ``Hilbert space'' or Fock module for~$W[V]$.

It is a remarkable fact, suggested by physicists, that there is also a map
\[
Y: {\rm Vac}[V] \to \End({\rm Vac}[V])[[z, z^{-1}]],
\]
known as the {\em vertex operator} or {\em state-field correspondence},
which is closely related to the $W[V]$-action.
The intuition behind $Y$ is that for any map $\phi: \CC^\times \to V$ and for any point $w \in \CC^\times$, 
there is a restriction of $\phi$ to a little disk around $w$.
That is, there is a restriction map $r_w: V[z,z^{-1}] \to V[[t]]$, where $t$ is the local coordinate $t = z-w$.
Hence there is a $w$-dependent map $\cO(V[[t]]) \to \cO(V[z,z^{-1}])$ by pulling back a function along $r_w$;
this map determines a $w$-dependent action of $\cO(V[[t]])$ on $\cO(V[[z]])$.
One can, in essence, quantize this map, by trying to extend the formulas of the $W[V]$-action,
and this leads to the map~$Y$.
If one axiomatizes the behavior of $Y$, one is led to the notion of a vertex algebra.

@@

When we try to quantize, we run into an interesting phenomenon:
the Hamiltonian action $H_\rho^{\rm loc}$ does not extend as a Lie algebra map into the Weyl algebra $W[V]$,
although it makes sense as a linear map.
The failure to respect the brackets determines, however, a cocycle on the loop algebra
by
\[
\alpha_V( x \otimes f(z) , y \otimes g(z)) = \Tr_V(xy) \Res(f \, \partial_z g).
\]
\owen{Check conventions.}
There is thus a central extension of the loop algebra 
\[
\CC c \to \widehat{\fg} \to \fg[z,z^{-1}]
\]
so that 
\[
[x \otimes f, y \otimes g] = [x,y] \otimes fg + \alpha_V( x \otimes f , y \otimes g) c,
\]
and there is a Lie algebra map
\[
\widehat{H}^{\rm loc}_\rho: \widehat{\fg} \to W[V],
\]
sometimes called {\em free field realization} of the affine Lie algebra~$\widehat{\fg}$.
This relationship --- and its extension to a map between the vertex algebras associated to $\widehat{\fg}$ and $V$ --- is a modest \owen{better adjective?} extract of the rich dialogue that has developed between complex analysis, representation theory, and the physics of chiral conformal field theory.
(To get Riemann surfaces into the game, we will need a richer framework than ordinary algebra.)

@@

There are natural central extensions of this sphere algebra as {\em $L_\infty$ algebras},
parallel to the affine Lie algebras appearing as central extensions of the loop algebra.
An $L_\infty$ algebra is a generalization of a Lie algebra;
it is graded vector space $L$ equipped with a countable collection of ``higher brackets'' $[-]_k: L^{\otimes k} \to L$ that satisfy a system of equations generalizing the Jacobi condition for a standard Lie algebra.
For an overview (and an actual definition!) we recommend \owen{add citations}.

In our setting of $\fg^\bullet_d$, Faonte-Hennion-Kapranov introduced central extensions that only affect the $d$-bracket.
For any $\fg$-invariant degree $d+1$ polynomial  $\theta \in \Sym^{d+1}(\fg^*)^\fg$,  the FHK  cocycle is
\[
\label{fhk cocycle}
\begin{array}{cccc}
\theta_{\rm FHK} : & (A_d \tensor \fg)^{\tensor (d+1)} & \to & \CC\\ 
& a_0 \otimes \cdots \otimes a_d & \mapsto & \Res_{z=0} \theta(a_0,\partial a_1,\ldots,\partial a_d)
\end{array}.
\]
This cocycle has cohomological degree $2$ and so determines an unshifted central extension as $L_\infty$ algebras
\[
\CC c \to \widetilde{\fg}^\bullet_{d, \theta} \to \fg^\bullet_d ,
\]
so that the usual bracket (the 2-bracket) of $\fg^\bullet_d$ is unchanged but there is now a nontrivial $d$-bracket defined by~$\theta_{\rm FHK}$.
They dubbed these the {\em higher Kac-Moody algebras}, 
as they specialize to the affine Lie algebras when~$d =1$.

These $L_\infty$ algebras are not just introduced on a whim.
In \cite{FHK} they are shown to appear naturally in various contexts,
notably in studying the derived moduli of $G$-bundles on a complex varieties of dimension~$d$.
\owen{Other applications?}

In addition to the higher Kac-Moody algebras, there are also natural higher analogs of the Weyl algebra of the loop space.
These appear as the observables for a holomorphic field theory known as a $\beta\gamma$ system.

Let $V$ be a symplectic complex vector space and consider the formal sphere algebra $A^{\wedge}_d \otimes V$,
where $A^{\wedge}_d$ denotes the completion of $A_d$ (i.e., where polynomials in $z_j$ and $z_j^*$ are replaced by formal power series).
\owen{Should we introduce a pithy notation?}
Then there is a dg commutative algebra of functions on this space,
which we denote $\cO(A^{\wedge}_d \otimes  V)$,
and it has a kind of Poisson bracket, arising from the residue pairing on $A^{\wedge}_d$ and the symplectic form on $V$,
just as $\cO(V((z)))$ has a Poisson bracket.
It will not surprise the reader to learn that there is a natural deformation quantization $W_d[V]$ into a dg Weyl algebra.
A physicist would interpret this algebra as the canonical and radial quantization of the theory on~$\CC^d - \{0\}$.

If $V$ is a representation of a Lie algebra $\fg$,
there is a local symmetry 
\[
\fg^\bullet_d \to \cO(A^{\wedge}_d \otimes V),
\]
just as $\fg[z,z^{-1}]$ acts on $\cO(V((z)))$. 
One can ask if this action lifts to a quantum symmetry.
The obstruction is, remarkably, an FHK cocycle:
\[
\alpha_{V,d}(x_1 \otimes f_1, \ldots, x_{d+1} \otimes f_{d+1}) = \Tr_V( x_1 \cdots x_{d+1}) \Res(f_1 \partial f_2 \cdots \partial f_{d+1}).
\]
\owen{Fix conventions}
In this sense, the higher Kac-Moody algebras admit a higher free field realization,
but a richer story is possible once we have a higher version of vertex algebras.
We will return to this topic once we have the rich language of factorization algebras.
